---
title: "A Guide to the BKP R package"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
```


## Installation

You can install the stable version of **BKP** from [CRAN](https://CRAN.R-project.org/package=BKP) with:

```{r, eval=FALSE}
install.packages("BKP")
```

Or install the development version from [GitHub](https://github.com/Jiangyan-Zhao/BKP) with:

```{r, eval=FALSE}
# install.packages("pak")
pak::pak("Jiangyan-Zhao/BKP")
```

## BKP model

```{r setup}
library(BKP)
library(tgp)
```


### Example 1

Let $x\in[-2,2]$, and suppose the true Bernoulli probability function is given by
$$
\pi_{1}(x) = \frac{1}{1+e^{-3x}}.
$$

```{r, fig.width = 6, fig.height = 6}
# Define true success probability function
true_pi_fun <- function(x) {
  1/(1+exp(-3*x))
}
X <- seq(-2, 2, length = 1000)
true_pi <- true_pi_fun(X)
plot(X, true_pi, type = "l", lwd = 2, xlab = "x", ylab = "Probability")
```

We aim to fit the BKP model based on seven input locations that are uniformly distributed over $[-2,2]$, with each location associated with a binomial observation having a maximum trial count of 100.

```{r, fig.width = 6, fig.height = 6}
set.seed(123)
# Data generation
n <- 7
Xbounds <- matrix(c(-2,2), nrow = 1)
X <- lhs(n = n, rect = Xbounds)
true_pi <- true_pi_fun(X)
m <- sample(100, n, replace = TRUE)
y <- rbinom(n, size = m, prob = true_pi)

# Fit BKP model
BKP_model_1D_1 <- fit.BKP(X, y, m, Xbounds = Xbounds)

# Plot results
plot(BKP_model_1D_1)

# Add true probability function
Xnew = matrix(seq(-2, 2, length = 100), ncol=1)
true_pi <- true_pi_fun(Xnew)
lines(Xnew,true_pi, col = "black", lwd = 2)
legend(x = -2.24,y = 0.89, lwd = 2, bty = "n", col = "black",
       legend = "Ture Probability", inset = 0.02)
```


### Example 2

The first example is essentially a generalized linear model with a smooth logit link, and thus poses limited modeling complexity. To demonstrate the capability of the BKP model in handling more challenging classification structures, we consider a second example with a highly nonlinear underlying probability surface. Define the true Bernoulli probability as
$$
\pi_{2}(x) = \frac{1}{2}\left[ 1+e^{-x^{2}}\cos \left( 10\frac{1-e^{-x}}{1+e^{-x}} \right) \right],
$$
where $x \in [-2,2]$. This example involves rapid local oscillations and strong nonlinearity, making it substantially more difficult to fit than Example 1. Here, we increase the number of locations to 30.

```{r, fig.width = 6, fig.height = 6}
# Define true success probability function
true_pi_fun <- function(x) {
  (1 + exp(-x^2) * cos(10 * (1 - exp(-x)) / (1 + exp(-x)))) / 2
}
X <- seq(-2, 2, length = 1000)
true_pi <- true_pi_fun(X)
plot(X, true_pi, type = "l", lwd = 2,
     xlab = "x", ylab = "Probability")
```


```{r, fig.width = 6, fig.height = 6}
set.seed(123)
# Data generation
n <- 30
Xbounds <- matrix(c(-2,2), nrow = 1)
X <- lhs(n = n, rect = Xbounds)
true_pi <- true_pi_fun(X)
m <- sample(100, n, replace = TRUE)
y <- rbinom(n, size = m, prob = true_pi)

# Fit BKP model
BKP_model_1D_2 <- fit.BKP(X, y, m, Xbounds = Xbounds)

# Plot results
plot(BKP_model_1D_2)

# Add true probability function
Xnew = matrix(seq(-2, 2, length = 100), ncol=1)
true_pi <- true_pi_fun(Xnew)
lines(Xnew,true_pi, col = "black", lwd = 2)
legend(x = -2.24,y = 0.89, lwd = 2, bty = "n", col = "black",
       legend = "Ture Probability", inset = 0.02)
```

### Example 3

Let $\vec{x} \in [0,1]^2$, and define the latent surface using a re-scaled version of the Goldstein–Price function:
$$
\begin{aligned}
	f(\vec{x})=& \frac{\log[\{1+a(\vec{x})\}\{30+b(\vec{x})\}-8.6928]}{2.4269}, \quad \text{with}\\
	a(\vec{x}) =& \left(4 x_1 + 4 x_2 - 3 \right)^2  \times  \notag  \\
	& \{75 - 56 \left(x_1 + x_2 \right) + 3\left(4 x_1 - 2 \right)^2 + 6\left(4 x_1 - 2 \right)\left(4 x_2 - 2 \right) + 3\left(4 x_2 - 2 \right)^2\}, \notag \\
	b(\vec{x}) =& \left(8 x_1 - 12 x_2 +2 \right)^2 \times \notag \\
	& \{-14 - 128 x_1 + 12\left(4 x_1 - 2 \right)^2 + 192 x_2 - 36\left(4 x_1 - 2 \right)\left(4 x_2 - 2 \right) + 27\left(4 x_2 - 2 \right)^2 \}.\notag 
\end{aligned}
$$

The true Bernoulli probability surface is then defined by 
$$
	\pi_3(\vec{x}) = \Phi\{f(\vec{x})\}, 
$$
where $\Phi(\cdot)$ is the cumulative distribution function of the standard normal distribution. 

```{r, fig.width = 6.5, fig.height = 6}
# Define 2D latent function and probability transformation
true_pi_fun <- function(X) {
  if(is.null(nrow(X))) X <- matrix(X, nrow=1)
  m <- 8.6928
  s <- 2.4269
  x1 <- 4*X[,1]- 2
  x2 <- 4*X[,2]- 2
  a <- 1 + (x1 + x2 + 1)^2 *
    (19- 14*x1 + 3*x1^2- 14*x2 + 6*x1*x2 + 3*x2^2)
  b <- 30 + (2*x1- 3*x2)^2 *
    (18- 32*x1 + 12*x1^2 + 48*x2- 36*x1*x2 + 27*x2^2)
  f <- log(a*b)
  f <- (f- m)/s
  return(pnorm(f))  # Transform to probability
}
x1 <- seq(0, 1, length.out = 100)
x2 <- seq(0, 1, length.out = 100)
X <- expand.grid(x1 = x1, x2 = x2)
true_pi <- true_pi_fun(X)
df <- data.frame(x1 = X$x1, x2 = X$x2, True = true_pi)
print(BKP:::my_2D_plot_fun("True", title = "True Probability", data = df))
```

To construct the training data, we generate a LHD of size $100$ over $[0,1]^2$. Each location is associated with a binomial observation whose number of trials is randomly drawn from $\{1, \dots, 100\}$.

```{r, fig.width = 9, fig.height = 8}
set.seed(123)
# Data generation
n <- 100
Xbounds <- matrix(c(0, 0, 1, 1), nrow = 2)
X <- lhs(n = n, rect = Xbounds)
true_pi <- true_pi_fun(X)
m <- sample(100, n, replace = TRUE)
y <- rbinom(n, size = m, prob = true_pi)

# Fit BKP model
BKP_model_2D <- fit.BKP(X, y, m, Xbounds=Xbounds)

# Plot results
plot(BKP_model_2D)
```

### Example 4

We next consider a binary classification task using the *Two Spirals* dataset, a well-known benchmark consisting of two intertwined spirals in a bounded two-dimensional input space. 

We generate $n = 200$ observations using the `mlbench.spirals` function from the R package [mlbench](https://CRAN.R-project.org/package=mlbench), with two complete rotations and additive Gaussian noise of standard deviation `sd = 0.05`. The inputs $\vec{x}$ are constrained to the domain $[-1.7, 1.7]^2$, and the binary class labels are encoded as 0 and 1. We fit the BKP model using a fixed prior specification with `r_0 = 0.1` and `p_0 = 0.5`.

```{r, fig.width = 13, fig.height = 6}
library(mlbench) 
set.seed(123)

# Data
n <- 200
data <- mlbench.spirals(n, cycles = 2, sd = 0.05)
X <- data$x
y <- as.numeric(data$classes) - 1  # Convert to 0/1 for BKP
m <- rep(1, n)
Xbounds <- rbind(c(-1.7, 1.7), c(-1.7, 1.7))

# Fit model
BKP_model_Class <- fit.BKP(
  X, y, m, Xbounds = Xbounds,
  prior = "fixed", r0 = 0.1, loss = "log_loss")

# Plot results
plot(BKP_model_Class)
```
